### 2014-04-22 rscripts/manuscript.Rnw
eh, commenting out a thing

Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  They watch the watchers as well, and increasing effort to reveal them should be expected to encourage adaptation.  Yet, as impossible as it is for outsiders to perfectly see the covert group, it is equally impossible for these groups to perfectly hide and accomplish anything of note.

>>>>

% Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  They watch the watchers as well, and increasing effort to reveal them should be expected to encourage adaptation.  Yet, as impossible as it is for outsiders to perfectly see the covert group, it is equally impossible for these groups to perfectly hide and accomplish anything of note.

### 2014-04-22 rscripts/manuscript.tex
eh, commenting out a thing

\usepackage{Sweave}
\begin{document}
\input{manuscript-concordance}
\begin{flushleft}
{\Large
\textbf{Modeling Requirements for Simulating Covert Social Groups}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carl A. B. Pearson$^{1,\ast}$, 
Edo Airoldi$^{2}$,
Edward Kao$^{2}$,
Burton Singer$^{1}$, 
\\
\bf{1} Emerging Pathogens Institute, University of Florida, Gainesville, FL, USA
\\
\bf{2} Statistics, Harvard University, Cambridge, MA, USA
\\
$\ast$ E-mail: cap10 \{at\} ufl \{dot\} edu
\end{flushleft}
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  Moreover, increasing active observation would be expected to perturb these social systems.  However, these groups never have perfect cover.  Their secrecy may make them conspicuously absent from the everyday activity of the background population.  Finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc. -- that might predict their actions.

The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.  With any covert network, there are strong constraints on passively and actively obtainable direct empirical data, overwhelming amounts of indirect data, and competing mechanical explanations for what is observed.  Studying the detailed history of any one network, or set of them, will not necessarily provide insight about how to effectively detect such networks in the future.  This situation recommends simulation of a variety of scenarios with accompanying detection strategies. Because of the censoring issues in the data, simulation requires particular attention to characterizing uncertainty and avoiding overfitting.

Herein, we discuss these issues in terms of network models.  Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) some philosophy relative to implementation.

As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.  We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g. Receiver Operator Characteristic).  Finally, we also demonstrate a small model extensionand show how it affects the (unextended) detection performance.

%\todoCP{{\em I think not for this round, but maybe:} Finally, we consider the implications of {\em forged} messages.  In the basic model, we consider incomplete information about the communications, but the available information is always accurate.  In this extension, we allow the Observer and the clandestine group to forge messages.  We again measure various Observer performance traits relative to properties of the observed network.}

@@

\section*{Introduction}
For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior groups operating at the edge of the observable is a principle concern.  In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.  Of course the revealation of these activities has broadly sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.  Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.

The underlying drive for these opposed efforts is the implacably expanding byte trail.  Once, the recorded information of an entire life might have amounted only to a few bytes -- e.g., parish records on births and deaths, perhaps including morbidity.  Now information era tools produce a near endless supply of 1s and 0s.  In some cases, this production is a side effect of organized commercial instinct -- Google monetizing search through targeted advertisement, for example.  In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.  However, in many cases this bitwise production is an engineering consequence of the associated technology.  E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers with almost no physical exchange, and of course any use of the internet produces veritable bit contrails.  This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.

Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.  We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.  Like ``algorithmic'' bubble prediction, we side more with Fama than Shiller.

Many such filters are based on social network descriptions.  These social networks can be a compact way to represent measured relations in a population.  However, these approaches all boil down to measuring properties of mathematical representation, with those properties having a tenuous connection to a mechanistic explanation of those phenomena.

>>>>

\usepackage{Sweave}
\begin{document}
\input{manuscript-concordance}
\begin{flushleft}
{\Large
\textbf{Agent-Based Modeling of Covert Social Groups, Applying Software Engineering Patterns and Planning for Cluster Computation}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carl A. B. Pearson$^{1,\ast}$, 
Edo Airoldi$^{2}$,
Edward Kao$^{2}$,
Burton Singer$^{1}$, 
\\
\bf{1} Emerging Pathogens Institute, University of Florida, Gainesville, FL, USA
\\
\bf{2} Statistics, Harvard University, Cambridge, MA, USA
\\
$\ast$ E-mail: cap10 \{at\} ufl \{dot\} edu
\end{flushleft}
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We discuss covert social network analysis (SNA) in the larger scientific context, highlight pitfalls, and propose what we view as necessary analysis that seems absent.  Because SNA typically occurs on aggregated observations (e.g., ties are across types and times), and furthermore covert SNA involves at least some (and often severe) censoring of those observations, it is critical that the implications of that aggregation and observation process be considered.

We demonstrate an implementation framework for performing that kind of analysis, based on a distributed, agent-based model.  We also apply past software engineering insights to show how they can make for more natural (and therefore comprehensible and compartmentally testable) agent-based models in this framework.

We close with discussion of how this framework might be used in a practical application, and how it is might be pieced together with other analytical tools.

@@

Covert social groups are difficult to observe directly, by definition: members either actively thwart discovery or leverage passive features of their surrounding population to avoid detection.  They watch the watchers as well, and increasing effort to reveal them should be expected to encourage adaptation.  Yet, as impossible as it is for outsiders to perfectly see the covert group, it is equally impossible for these groups to perfectly hide and accomplish anything of note.

% Their secrecy may make them conspicuously absent from the everyday activity of the background population.  Finally there are potential mechanical charactericterizations -- their composition, goals, tactics, etc. -- that might predict their actions.
% 
% Often these mechanically characterizations are framed in terms of social networks.  Individuals and their ties are thought to have certain graph theoretic features based on their importance, and the observation of these features is then used to focus attention accordingly.  Yet, often these networks are little more than diverse shadows on the wall.  They are not the mechanisms or the phenomena, they are not even observation: they are mere representation.
% 
% The structure of covert networks is highly variable from one context to another, and they change composition and interaction patterns over time.  With any covert network, there are strong constraints on passively and actively obtainable direct empirical data, overwhelming amounts of indirect data, and competing mechanical explanations for what is observed.  Studying the detailed history of any one network, or set of them, will not necessarily provide insight about how to effectively detect such networks in the future.  This situation recommends simulation of a variety of scenarios with accompanying detection strategies. Because of the censoring issues in the data, simulation requires particular attention to characterizing uncertainty and avoiding overfitting.
% 
% Herein, we discuss these issues in terms of network models.  Specifically, we review (i) incorporating observation uncertainty, (ii) the need for both back- and foreground populations, (iii) avoiding over-fitting, and (iv) some philosophy relative to implementation.
% 
% As part of that philosophical discussion, we walkthrough the exercise of procedurally generating back- and foreground populations, simulating communication among those individuals, and filtering those communications via an observation model.  We then address the issues of (i) fitting that model against data, and (ii) analyzing performance of the opposing sides (e.g. Receiver Operator Characteristic).  Finally, we also demonstrate a small model extensionand show how it affects the (unextended) detection performance.

\section*{Introduction}
For investigators ranging from anthropologists to law enforcement, the desire to find, study, and modify the behavior of groups operating at the edge of the observable is a principle concern.  In the meta-theatre of the ironic, various media outlets have responded to public interest by applying this lens to the activities of intelligence organizations world-wide -- themselves desperate to undertake this very task against other clandestine actors.  The revealation of these activities has sparked the symmetric concern: being able to operate clandestinely in an age of ubiquitous monitoring.  Criminal organizations have long appreciated the value of operating in secret, as have groups subject to State-sponsored abuse or industrial espionage, and time will tell whether or not this spark catches fire in general behavior.

The underlying drive for these opposed efforts is the implacably expanding byte trail.  Once, the recorded information of an entire life might have amounted only to a few bytes -- e.g., parish records on births and deaths, perhaps including morbidity.  Now information era tools produce a near endless supply of 1s and 0s.  In some cases, this production is a side effect of organized commercial instinct -- Google monetizing search through targeted advertisement, for example.  In other cases, individuals are voluntarily advancing their own brand -- Instagram ``selfies'', e.g.  However, in many cases this bitwise production is an engineering consequence of the associated technology.  E.g., cellular phones transmit constant location data, financial systems are increasing just digital ledgers substituting the certainty of physical exchange with sophisticated book-keeping to ensure accounts are square, and of course any use of the internet produces veritable bit contrails.  This rate of production far exceeds the personal processing capability of any practically-sized team of analysts.

Hence, these teams employ computer-based, heuristic filtration to decide which data to record, to review, and to obtain.  We avoid saying ``algorithmic'' at this point: ``algorithmic'' implies strong, logically inevitable conclusions from inputs.  Like ``algorithmic'' economic bubble prediction, we side more with Fama than Shiller.

Many such filters are based on social network descriptions.  Social networks are a compact way to represent measured relations in a population.  That compression, however, often involves irreversible aggregations: in strength -- e.g., co-employment as measured by working at the same university versus the same department versus in same the laboratory; in type -- e.g., engaging in commerce versus attending the same church versus having married relations; and in time -- e.g., co-appearance at a coffeeshop months ago versus attending classes together on a weekly schedule versus living together.

The mathematics may provide a formally guaranteed result from some particular input, but the more aggregation that occurs translating observations -- themselves imperfect snapshots of actual phenomena -- to representation, the more convincing evidence that must be provided before we should accept that those representations (and subsequent results from manipulating them) have meaning.
